---
title: "Ferry Writeup"
author: "From your friends at SLED analytics"
---


We have long been interested in WSDOT ferry on-time rates and finally decided to file a Public Records Request in an attempt to summarize this data. Here we present findings for on-time rates for calendar year 2014.

##On time departures
WSDOT uses GPS trackers on each ferry and is therefore able records departure times to the second, so initially we looked at the difference in scheduled departure time versus actual departure time. Because WSDOT runs a [holiday schedule](http://www.wsdot.wa.gov/Ferries/infodesk/faq/general_info/) on most Federal holidays, we excluded those specific dates.

```{r, echo=FALSE, fig.width=6}
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)

`%notin%` <- function(x,y) !(x %in% y)

raw <- read.csv("data/PDR-15-1195.csv", colClasses="character")
raw$scheduled_departure_dt <- as.POSIXlt(raw$scheduled_departure)
raw$scheduled_departure <- as.Date(raw$scheduled_departure)
raw$actual_departure_dt <- as.POSIXlt(raw$actual_departure)
raw$scheduled_hour <- raw$scheduled_departure_dt$hour
raw$difference <- as.numeric(difftime(raw$actual_departure, raw$scheduled_departure_dt, units = "min"))
raw$departure_date <- as.Date(raw$scheduled_departure)

#classify days in raw data set
raw$wday <- raw$scheduled_departure_dt$wday
raw$day <- "Weekday"
raw$day[raw$wday == 0 | raw$wday == 6] <- "Weekend"

holidays <- as.Date(c("2014-01-01", "2014-01-20", "2014-02-17",
                      "2014-05-26", "2014-07-04", "2014-09-01",
                      "2014-10-13", "2014-11-11",
                      "2014-11-27", "2014-12-25"))
raw$day[raw$departure_date %in% holidays] <- "Holiday"

raw$delay <- abs(raw$difference)
raw$delay[raw$difference < 0] <- NA

all.days <-
  raw %>%
  #subset(scheduled_departure %notin% holidays) %>%
  select(scheduled_departure, route_name, difference) %>%
  group_by(scheduled_departure, route_name) %>%
  summarise_each(funs(mean), difference)

ggplot(all.days, aes(scheduled_departure, difference)) +
  geom_point() + stat_smooth()
```

Each point represents the mean departure delay for that date for all ferries with a blue trendline overlaid. As a whole, the average daily delay in ferries was 3.5 minutes in 2014.

##On time departures by route
```{r, echo=FALSE, fig.height=8, fig.width=7}
ggplot(all.days, aes(scheduled_departure, difference, group=route_name, color=route_name)) +
  geom_path(alpha = 0.9)
```

Then we focused only on ferries that sail the lower Puget Sound (ie non-Orcas Islands ferries)

```{r, echo=FALSE, fig.height=8, fig.width=10}
pugetSound <- c("Fauntleroy - Southworth",
                "Fauntleroy - Vashon",
                "Mukilteo - Clinton",
                "Pt. Defiance - Tahlequah",
                "Seattle - Bainbridge Island",
                "Seattle - Bremerton",
                "Southworth - Vashon",
                "Keystone - Port Townsend")

pugetSoundRoutes <- 
  raw %>%
  subset(delay >= 0 & route_name %in% pugetSound) 

hour.levels <- rev(0:23)
pugetSoundRoutes$scheduled_hour <- factor(pugetSoundRoutes$scheduled_hour, levels = 23:0, ordered=T)
names(pugetSoundRoutes)[14:15] <- c("Day", "Delay")

ggplot(pugetSoundRoutes) + theme_bw() +
  geom_point(aes(departure_date, scheduled_hour, size = Delay, col = Day)) +
  facet_wrap(~route_name, scales="free") + 
  scale_size_continuous(range = c(.1, 6)) + 
  ggtitle("Delays for All 2014 Puget Sound Sailings")+
  scale_x_date(labels = scales::date_format("%b")) +
  xlab("Departure Date") + ylab("Scheduled Hour of Departure") +
  scale_y_discrete(breaks = c(0, 5, 9, 13, 17, 20, 23), labels = c("Midnight", "5 am", "9 am", "1 pm", "5 pm", "8pm", "11pm"))
  
```


What about looking at the individual boats themselves?
```{r, echo=FALSE, fig.height=8, fig.width=10}

cleaned <- raw[abs(raw$difference) < 121, ]
cleaned$week <- lubridate::week(cleaned$departure_date)
weekly <- do.call('rbind', by(cleaned, interaction(cleaned$week, cleaned$actual_vessel_name), function(x) {
  data.frame("min" = min(x$difference), 
             "max" = max(x$difference), 
             "median" = median(x$difference),
             "week" = x$week[1],
             "vessel" = x$actual_vessel_name[1])
}))

wlong <- reshape2::melt(weekly, id.vars = c("week", "vessel"))
ggplot(wlong) + theme_bw() + 
  geom_line(aes(x = week, y = value, linetype = variable)) + 
  facet_wrap(~vessel, scales="free") + 
  ggtitle("Difference from Scheduled to Actual Departure WSDOT Ferries 2014\nWeekly Min, Max, and Median - Boxed by Departure Terminal") +
  xlab("Week of Year") + ylab("Actual Departure Minutes from Scheduled Departure") + 
  scale_linetype_manual(name = "Weekly Summaries", values = c("dotted", "dashed", "solid"))

```

Finally, lets look at the quantiles
```{r, echo=FALSE, fig.height=8, fig.width=10}
medians <- do.call('rbind', by(cleaned, cleaned$route_name, function(x) {
  data.frame("q05" = as.numeric(quantile(x$difference, .05)), 
             "q25"= as.numeric(quantile(x$difference, .25)),
             "q50" = median(x$difference),
             "q75" = as.numeric(quantile(x$difference, .75)), 
             "q95" = as.numeric(quantile(x$difference, .95)),
             "mean" = mean(x$difference), 
             "route" = x$route_name[1])
}))
medians <- arrange(medians, mean)
medians$route <- factor(medians$route, levels = medians$route)
mlong <- reshape2::melt(medians, id.vars = "route")
ggplot(mlong[mlong$variable != "mean", ]) + theme_bw() + 
  geom_point(aes(x = value, y = route, shape = variable, col = variable)) +
  scale_colour_discrete(name = "Summary", labels = c("5% quantile", "25% quantile", "median", "75% quantile", "95% quantile")) +
  scale_shape_discrete(name = "Summary", labels = c("5% quantile", "25% quantile", "median", "75% quantile", "95% quantile")) +
  ggtitle("WSDOT Ferries 2014 Scheduled vs Actual Departure Time\nSorted by Average Delay Time") + 
  xlab("Minutes Difference between Scheduled and Actual Departure") +
  ylab("Route")
```

