---
title: "Ferry Writeup"
author: "From your friends at SLED analytics"
---


I (we?) have long been interested in WSDOT ferry data and finally decided to act on that by filing a Public Records Request for raw data for all of 2014 and writing up a brief analysis for Seattle Transit Blog.

##On time departures
WSDOT records departure times to the second, so the initial analysis is to simply look at the difference in scheduled departure time versus actual departure time. WSDOT runs a [holiday schedule](http://www.wsdot.wa.gov/Ferries/infodesk/faq/general_info/) on most Federal holidays, so we excluded those dates.

```{r, echo=FALSE, fig.width=6}
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)

`%notin%` <- function(x,y) !(x %in% y)

raw <- read.csv("data/PDR-15-1195.csv", colClasses="character")
raw$scheduled_departure_dt <- as.POSIXlt(raw$scheduled_departure)
raw$actual_departure <- as.POSIXlt(raw$actual_departure)
raw$difference <- abs(as.numeric(raw$actual_departure - raw$scheduled_departure_dt) / 60)
raw$scheduled_departure <- as.Date(raw$scheduled_departure)

holidays <- as.Date(c("2014-01-01", "2014-05-26",
                      "2014-07-04", "2014-09-01",
                      "2014-11-27", "2014-11-25"))
regular.days <-
  raw %>%
  subset(scheduled_departure %notin% holidays) %>%
  select(scheduled_departure, route_name, difference) %>%
  group_by(scheduled_departure, route_name) %>%
  summarise_each(funs(mean), difference)

ggplot(regular.days, aes(scheduled_departure, difference)) +
  geom_point() + stat_smooth()
```

The blue line is a trendline showing the mean delay in departure for all ferries which was roughly 3 minutes per ferry in 2014.

##On time departures by route
```{r, echo=FALSE, fig.height=8, fig.width=7}
ggplot(regular.days, aes(scheduled_departure, difference, group=route_name, color=route_name)) +
  geom_path(alpha = 0.9)
```

