---
title: "Ferry Curious or How I learned to stop worrying and accept WSDOT"
author: "From your friends at SLED analytics"
output: html_document
---

We have long been interested in WSDOT ferry data, specifically on-time rates, and finally decided to file a Public Records Request in an attempt to gain access to this data. Here we present findings for on-time rates for calendar year 2014.

First a few caveats: Because WSDOT runs a [holiday schedule](http://www.wsdot.wa.gov/Ferries/infodesk/faq/general_info/) on most Federal holidays, we excluded those specific dates. Also, for the purposes of brevity for this report we solely focused on the routes in Central Puget Sound (non-Orcas Islands routes).

##On time departures
WSDOT uses GPS trackers on each ferry and is therefore able to record arrival and departure times to the second, so we initially looked at the difference in scheduled departure time versus actual departure time.

```{r, echo=FALSE, fig.width=6}
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)

`%notin%` <- function(x,y) !(x %in% y)

raw <- read.csv("data/PDR-15-1195.csv", colClasses="character")
raw$scheduled_departure_dt <- as.POSIXlt(raw$scheduled_departure)
raw$scheduled_departure <- as.Date(raw$scheduled_departure)
raw$actual_departure_dt <- as.POSIXlt(raw$actual_departure)
raw$scheduled_hour <- raw$scheduled_departure_dt$hour
raw$difference <- as.numeric(difftime(raw$actual_departure, raw$scheduled_departure_dt, units = "min"))
raw$departure_date <- as.Date(raw$scheduled_departure)

#classify days in raw data set
raw$wday <- raw$scheduled_departure_dt$wday
raw$day <- "Weekday"
raw$day[raw$wday == 0 | raw$wday == 6] <- "Weekend"

holidays <- as.Date(c("2014-01-01", "2014-01-20", "2014-02-17",
                      "2014-05-26", "2014-07-04", "2014-09-01",
                      "2014-10-13", "2014-11-11",
                      "2014-11-27", "2014-12-25"))
raw$day[raw$departure_date %in% holidays] <- "Holiday"

raw$delay <- abs(raw$difference)
raw$delay[raw$difference < 0] <- NA

pugetSound <- c("Fauntleroy - Southworth",
                "Fauntleroy - Vashon",
                "Mukilteo - Clinton",
                "Pt. Defiance - Tahlequah",
                "Seattle - Bainbridge Island",
                "Seattle - Bremerton",
                "Southworth - Vashon",
                "Keystone - Port Townsend")

all.days <-
  raw %>%
  subset(route_name %in% pugetSound) %>%
  select(scheduled_departure, route_name, difference) %>%
  group_by(scheduled_departure, route_name) %>%
  summarise_each(funs(mean), difference)

ggplot(all.days, aes(scheduled_departure, difference)) +
  geom_point(size=0.3) + stat_smooth()
```

Each point represents the mean departure delay on that date for all ferries with a trendline (blue) overlaid. As a whole, the data shows variation in departure time is kept pretty minimal with a large bump in departure time variation towards the end of summer. Average daily depature delay in ferries was approximately 3 minutes for all ferry routes of interest.


```{r, echo=FALSE}
ggplot(all.days, aes(scheduled_departure, difference)) + geom_point(size=0.3) + stat_smooth() +
  scale_x_date(labels = scales::date_format("%b")) +
  facet_wrap(~route_name)
```

The above figure shows the average daily departure delay for the eight routes of interest over the course of the year. The seasonal trend present in Figure 1 is visible in almost all of these routes as well.


Next, we focused only on time delays over the course of the day while accounting for types of days (weekdays, weekends or holidays).
```{r, echo=FALSE}


pugetSoundRoutes <- 
  raw %>%
  subset(delay >= 0 & route_name %in% pugetSound) 

hour.levels <- rev(0:23)
pugetSoundRoutes$scheduled_hour <- factor(pugetSoundRoutes$scheduled_hour, levels = 23:0, ordered=T)
names(pugetSoundRoutes)[14:15] <- c("Day", "Delay")

ggplot(pugetSoundRoutes) + 
  geom_point(aes(departure_date, scheduled_hour, size = Delay, col = Day)) +
  facet_wrap(~route_name) + 
  scale_size_continuous(range = c(.1, 6)) + 
  ggtitle("Delays for All 2014 Puget Sound Sailings")+
  scale_x_date(labels = scales::date_format("%b")) +
  xlab("Departure Date") + ylab("Scheduled Hour of Departure") +
  scale_y_discrete(breaks = c(0, 5, 9, 13, 17, 20, 23), labels = c("Midnight", "5 am", "9 am", "1 pm", "5 pm", "8pm", "11pm"))
  
```

The above plot shows both number of minutes of delay colored by the type of day for every hour per day for each of the eight routes of interest.

Focus on only the Seattle-Bainbridge Island Route:
```{r, echo=FALSE}
bi.only <- subset(pugetSoundRoutes, route_name == "Seattle - Bainbridge Island")
ggplot(bi.only) + 
  geom_point(aes(departure_date, scheduled_hour, size = Delay, col = Day)) +
  facet_wrap(~route_name) + 
  scale_size_continuous(range = c(.1, 6)) + 
  ggtitle("Delays for All 2014 Seattle - Bainbridge Island Sailings")+
  scale_x_date(labels = scales::date_format("%b")) +
  xlab("Departure Date") + ylab("Scheduled Hour of Departure") +
  scale_y_discrete(breaks = c(0, 5, 9, 13, 17, 20, 23), labels = c("Midnight", "5 am", "9 am", "1 pm", "5 pm", "8pm", "11pm"))
```

Basically, try to avoid weekend afternoon sailings in the summer

WSDOT also tracks which boat is used on each route which allows for a look at boat efficiencies over time on a weekly basis
```{r, echo=FALSE}

cleaned <- 
  raw %>% 
  subset(abs(difference) < 121 & route_name %in% pugetSound)
                 
                 
cleaned$week <- lubridate::week(cleaned$departure_date)
weekly <- do.call('rbind', by(cleaned, interaction(cleaned$week, cleaned$actual_vessel_name), function(x) {
  data.frame("min" = min(x$difference), 
             "max" = max(x$difference), 
             "median" = median(x$difference),
             "week" = x$week[1],
             "vessel" = x$actual_vessel_name[1])
}))

wlong <- reshape2::melt(weekly, id.vars = c("week", "vessel"))
ggplot(wlong) + 
  geom_line(aes(x = week, y = value, linetype = variable)) + 
  facet_wrap(~vessel) + 
  ggtitle("Difference from Scheduled to Actual Departure WSDOT Ferries 2014\nWeekly Min, Max, and Median - Boxed by Departure Terminal") +
  xlab("Week of Year") + ylab("Actual Departure Minutes from Scheduled Departure") + 
  scale_linetype_manual(name = "Weekly Summaries", values = c("dotted", "dashed", "solid"))

```

Interesting to observe how often WSDOT swaps out different boats.


We have already looked at departure delays, what about looking at mean crossing times for each boat?
```{r, echo=FALSE}
raw <- 
  raw %>%
  subset(Actual_Arrival != "NULL")
raw$actual_arrival <- as.Date(raw$Actual_Arrival)
raw$actual_arrival_dt <- as.POSIXlt(raw$Actual_Arrival)
raw$crossing <- as.numeric(difftime(raw$actual_arrival_dt, raw$actual_departure_dt, units = "min"))
mean.times <-
  raw %>%
  subset(route_name %in% pugetSound) %>%
  select(scheduled_departure, route_name, crossing) %>%
  group_by(scheduled_departure, route_name) %>%
  summarise_each(funs(mean), crossing)

ggplot(mean.times, aes(scheduled_departure, crossing)) + geom_point(size=0.3) + stat_smooth() +
  scale_x_date(labels = scales::date_format("%b")) +
  facet_wrap(~route_name)
```

The above figures show that most ferries are very consistent after departure with the exception of the Fauntleroy - Southworth and Keystone - Port Townsend routes, which show a fair amount of variation. This is reflected in the following table showing standard deviation:

```{r, echo=FALSE}
var.table <- plyr::ddply(mean.times, "route_name", function(x) data.frame(variation=var(x$crossing))) 
knitr::kable(var.table)
```
Because these figures only look at crossing times, any variation in trip length is due to factors that must be addressed by the ferry captain.


Finally, lets look at the quantiles for the eight routes of interest sorted by average delay time
```{r, echo=FALSE}
medians <- do.call('rbind', by(cleaned, cleaned$route_name, function(x) {
  data.frame("q05" = as.numeric(quantile(x$difference, .05)), 
             "q25"= as.numeric(quantile(x$difference, .25)),
             "q50" = median(x$difference),
             "q75" = as.numeric(quantile(x$difference, .75)), 
             "q95" = as.numeric(quantile(x$difference, .95)),
             "mean" = mean(x$difference), 
             "route" = x$route_name[1])
}))
medians <- arrange(medians, mean)
medians$route <- factor(medians$route, levels = medians$route)
mlong <- reshape2::melt(medians, id.vars = "route")
ggplot(mlong[mlong$variable != "mean", ]) + 
  geom_point(aes(x = value, y = route, shape = variable, col = variable)) +
  scale_colour_discrete(name = "Summary", labels = c("5% quantile", "25% quantile", "median", "75% quantile", "95% quantile")) +
  scale_shape_discrete(name = "Summary", labels = c("5% quantile", "25% quantile", "median", "75% quantile", "95% quantile")) +
  ggtitle("WSDOT Ferries 2014 Scheduled vs Actual Departure Time\nSorted by Average Delay Time") + 
  xlab("Minutes Difference between Scheduled and Actual Departure") +
  ylab("Route")
```


One thing we would definitely be interested in would be the ferry capacity each day but that will have to wait for another analysis.